Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2022-10-03 text/Manuscript.Rmd
Figure update

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** (A) SEM image of a 3-day-old Platynereis larva.(B-D) Expression analysis of the NOS gene (magenta) using in situ HCR. (D) Co-stained image using acetylated α-tubulin antibody (ac α-tub: green). (B) Dorsal view of the larva at 3-day-old. (C and D) Anterior view of the larva at 2 (D) and 3-day-old (C). (E) Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia and axonal scaffold (ac α-tub: green). INNOSs with the entire cell body and axon labelled are shown with white arrowheads and with only the cell body labelled are shown with black arrowheads. Views of larvae at 3-day-old from the anterior side. (F and G) Five types of neurons (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a 3-day-old larva. The INNOS on the ventral side labelled with (E) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. (H) Wiring diagram of the cPRC circuit. <!---# This graph widget does not look very good. We need to edit this a bit more. --> ](../figures/Fig1.png){width="100%"}

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans [@Moroz2020]. To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Figure 3---figure supplement 1). Using previously published spatially mapped single-cell transcriptome data  [@Achim2015; @williams2017] to find candidate NIT-GCs involved in the cPRC circuit among 15 NIT-GCs, two NIT-GCs (NIT-GC1 and NIT-GC2) were found to be highly expressed in the cPRC (Figure 3---figure supplement 1). Expression analysis with c-opsin1 and MLD pNPs (pedal peptide neuropeptide precursor 2) [@Arendt2004; @williams2017], which have already been identified as cPRC markers using in situ HCR, showed that both NIT-GC1 and NIT-GC2 are strongly and specifically expressed in cPRC (Figure 3A and B, Figure 3---figure supplement 2). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Figure 3C and D). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). These results suggest that NIT-GC1 may function as a major target for NO.

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems [@Matsuda2017]. We co-expressed Green cGull and NIT-GC1 using the 2A self-cleaving peptide in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 (Figure 3---figure supplement 4). When the NO donor SNAP was applied to the co-expression cells, the fluorescence intensity in the cells was significantly increased compared to when NIT-GC1 was not expressed (Figure 3E and F). The increase in fluorescence intensity also did not occur when DMSO was applied (Figure 3G). To confirm whether this increase in fluorescence intensity is NIT domain-dependent response of NIT-GC1, mutant form of NIT-GC1 without the NIT domain was expressed and SNAP was applied (Figure 3---figure supplement 4A). As a result, the increase in fluorescence intensity was significantly reduced  (Figure 3H). These results indicate that NIT-GC1 is activated by NO to produce cGMP. They also show that the NIT domain is required for the activation of NIT-GC1.

@@

To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light. Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells. We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 30-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 30 min, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase occurs after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. To determine whether NO mediates such a feedback, we measured calcium signals in *NOS*-mutant larvae. While we detected the inhibition phase in cPRC in *NOS*-mutants, this was not followed by activation and calcium levels remained low. These results revealed an essential requirement for NOS signalling in the second phase of the cPRC response.

We next quantified the responses of the INNOS and INRGW interneurons following a UV/violet stimulus to the cPRCs. To unambiguously identify the cells from which we recorded calcium signals, we developed an on-slide immunostaining method. We used the cell-specific neuropeptide markers anti-RYamide (INNOS) and anti-RGWa (INRGW) [@Conzelmann2011] to mark the interneurons. Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 4---figure supplement 1). In both wild type and *NOS*-mutant larvae, INNOS cells showed increase calcium signals during the inhibition phase of the cPRCs. In contrast, calcium signals increased in INRGW cells in wild-type larvae during the second activation phase of cPRCs, a response that was lacking in *NOS*-mutants. This suggests that there is no circuit output to the Ser-h1 cells and the ciliary band in *NOS*-mutants, consistent with the observed behavioural defects.

**NO induces depolarisation of the cPRCs via NIT-GC1**

To determine the function of NO in the cPRC circuit, the intracellular dynamics of neurons in the cPRC circuit were investigated, focusing on NIT-GC1. First, we investigated the changes in NO during UV stimulation of the cPRC using DAF-FM, a fluorescent probe that can detect NO, in order to determine whether NO is produced by UV light. By injecting DAF-FM directly into zygotes, INNOS cells could be identified under live imaging. To visualise the location of cPRCs, the calcium indicator RGECO1a was also injected together, as cPRCs maintain a higher resting calcium levels compared to other cells. Direct UV stimulation to the cPRCs increased the fluorescence intensity of DAF-FM compared to stimulation of non-cPRC region (Fig. 5A). Next, we did calcium imaging of cells in the cPRC circuit during UV stimulation. We locally stimulated the cPRC cilia in wild-type with a 405 nm laser (UV) and detected a slight increase followed by a strong decrease in calcium levels (hyperpolarisation) (Fig. 5B). Turning off the UV after or during the decrease caused a strong increase (depolarisation) and a return to the same calcium level as before (Fig. 5B). These responses were similar to those reported previously [@Veraszto2018]. We performed a similar stimulation using NOS mutants, which hyperpolarised during the UV stimulation, but no depolarising responses were observed when UV was stopped (Fig. 5C). Similarly, in NIT-GC1 knockdown larvae using the morpholinos, no depolarisation after the hyperpolarisation response were observed (Fig. 5D). Interestingly, no hyperpolarising response during UV stimulation was found to occur in NIT-GC2 knockdown larvae. And we compared the changes of the INNOS calcium levels identified by their position in relation to cPRCs during UV stimulation in wild-type and in NOS mutants. In both cases, a depolarising response was observed at the same time as the hyperpolarising response of the cPRCs (Fig. 5E). Furthermore, to perform calcium imaging of INRGWs in UV stimulation, INRGWs were identified by immunostaining with RGWa antibodies after calcium imaging (Fig. 5F). It has been reported that the INRGW cells form many large synapses with synaptic vesicles on two serotonergic ciliomotor neurons (Ser-h1) that are known to play a role in the regulation of ciliary beating and the activity of the INRGW cells increased the negative correlation between the activity patterns of serotonergic neurons and the MC neuron (Williams et al., 2018). The results showed that depolarisation of INNOS occurs at the same time as hyperpolarisation of cPRC in the wild type. Identified INRGW cells were found to cause depolarisation at the same time as cPRC hyperpolarisation in the wild type (Fig. 5G). It was also found that in NOS mutants where cPRC hyperpolarisation does not occur, the INRGW depolarisation response also does not occur (Fig. 5H).

![**Figure 4. NO produced by UV stimulation to cPRCs induces a depolarising response in cPRC via NIT-GC1.** (A) The intensity of DAF-FM in INNOS changes over time when cPRC (top) and other cells (NC) (bottom) are stimulated with UV (405 nm). Hereafter, light purple squares indicate the timing of UV stimulation, gray traces show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark lines and the bands show the mean value and the standard error range of the mean. (B) Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in WT (purple) and NOS mutant (NOSΔ23) (gray) larvae. (C) Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in control (non-injection) and NIT-GC1 knockdown by morpholino (NITGC1-MO1 and 2) larvae. (D) Co-expression analysis image of the cPRC marker gene (NIT-GC1: magenta) and the INNOS marker gene (RYa-pNP: green) using in situ HCR. White arrowheads indicate INNOS strongly expressing RYa-pNP. (E) INNOS (blue dashed line) identified from calcium imaging using GCaMP6s by their position in relation to the cPRC (purple dashed line). (F) Comparison of changes over time in GCaMP6s intensities in cPRC (purple) and INNOS (blue) when cPRC is stimulated with UV in WT (left) and NOS mutant (NOSΔ11: right) larvae. (G) Comparison of changes over time in GCaMP6s intensities in cPRC (purple) and INRGW (green) when the cPRC are stimulated with UV in WT (left) and NOS mutant (NOSΔ23: right) larvae. (H)(I)](../figures/Fig4.png){width="100%"}

>>>>

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** (A) SEM image of a 3-day-old Platynereis larva.(B-D) Expression analysis of the NOS gene (magenta) using in situ HCR. (D) Co-stained image using acetylated α-tubulin antibody (ac α-tub: green). (B) Dorsal view of the larva at 3-day-old. (C and D) Anterior view of the larva at 2 (D) and 3-day-old (C). (E) Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia and axonal scaffold (ac α-tub: green). INNOSs with the entire cell body and axon labelled are shown with white arrowheads and with only the cell body labelled are shown with black arrowheads. Views of larvae at 3-day-old from the anterior side. (F and G) Five types of neurons (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a 3-day-old larva. The INNOS on the ventral side labelled with (E) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. (H) Wiring diagram of the cPRC circuit. <!---# This graph widget does not look very good. We need to edit this a bit more. -->](../figures/Fig1.png){width="100%"}

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans [@Moroz2020]. To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Figure 3---figure supplement 1). Using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017] to find candidate NIT-GCs involved in the cPRC circuit among 15 NIT-GCs, two NIT-GCs (NIT-GC1 and NIT-GC2) were found to be highly expressed in the cPRC (Figure 3---figure supplement 1). Expression analysis with c-opsin1 and MLD pNPs (pedal peptide neuropeptide precursor 2) [@Arendt2004; @williams2017], which have already been identified as cPRC markers using in situ HCR, showed that both NIT-GC1 and NIT-GC2 are strongly and specifically expressed in cPRC (Figure 3A and B, Figure 3---figure supplement 2). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Figure 3C and D). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). These two localisation differences suggest the existence of different intracellular signalling mechanisms in the ciliary and neuroaxonal regions within the cPRC.

**NIT-GC1 is required for the depolarising response of cPRC to UV stimulation**

To characterise the role of NIT-GCs within the cPRC during UV stimulation in more detail, the calcium sensor GCaMP6s were ubiquitously expressed in larvae in which NIT-GCs were knocked down using morpholino to examine changes in calcium levels in the cPRC. Local stimulation of the ciliated part of the cPRC with 405 nm light causes a transient decrease in calcium levels in the cPRC (hyperpolarisation), followed by an increase in calcium levels (depolarisation) when the light is switched off [@veraszto2018]. In NIT-GC1 morphant larvae, 405 nm light stimulation caused a decrease in calcium levels in the cPRC, but the increase in calcium levels that occurred after the light stimulation was stopped was strongly suppressed (Figure 3E). This result suggests that NIT-GC1 mediates depolarising responses after UV stimulation of cPRC and may be involved in extracellular signalling. Interestingly, in NIT-GC2 morphant larvae, an increase in calcium levels occurred immediately after light stimulation (Figure 3F). The different responses of cPRC upon knockdown of NIT-GC1 and NIT-GC2 suggest that there may be different intracellular signalling mechanisms associated with their respective localisations. We focused our analysis on NIT-GC1 because it was localised to cPRC neuroaxons within the neuroplexas region and is involved in extracellular signalling after UV stimulation. It is generally known that depolarisation in the PRC involves elevated cGMPs in the photoreceptors. We therefore used an in vitro system in which cultured cells expressed NIT-GC1 to assess the cGMP activity of NIT-GC1.

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems [@Matsuda2017]. We co-expressed Green cGull and NIT-GC1 using the 2A self-cleaving peptide in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 (Figure 3---figure supplement 4). When the NO donor SNAP was applied to the NIT-GC1 and Green cGull co-expression cells, the intracellular cGMP levels was significantly increased compared to when NIT-GC1 was not expressed (Figure 3G and H). The increase in fluorescence intensity also did not occur when DMSO was applied (Figure 3I). To confirm whether this increase in intracellular cGMP levels is NIT domain-dependent response of NIT-GC1, mutant form of NIT-GC1 without the NIT domain was expressed and SNAP was applied (Figure 3---figure supplement 4A). As a result, the increase in fluorescence intensity was significantly reduced (Figure 3J). These results indicate that NIT-GC1 produces cGMP via the NIT domain. In bacteria, it has been proposed that the NIT domain regulates cellular functions in response to changes in extracellular and intracellular nitrate and nitrite concentrations. Our results suggest that NIT-GC1 may be sensitive to NO. This could be a possible feedback pathway in which NIT-GC1 expressed in cPRC is activated by NO produced by IN-NOS. In order to further explore the role of NO in the cPRC circuit, we therefore compared calcium imaging of postsynaptic cells (IN-NOS, IN-RGW, Ser-h1, MC) in the cPRC circuit during UV stimulation of cPRC between WT and NOS mutant larvae.

@@

To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light (Figure 4---figure supplement 1). Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells. 
We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 20-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 20 sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase occurs after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. 
To determine whether NO mediates such a feedback, we measured calcium signals in *NOS*-mutant larvae. While we detected the inhibition phase in cPRC in *NOS*-mutants, this was not followed by activation and calcium levels remained low. These results revealed an essential requirement for NOS signalling in the second phase of the cPRC response.

We next quantified the responses of the INNOS and INRGW interneurons following a UV/violet stimulus to the cPRCs. To unambiguously identify the cells from which we recorded calcium signals, we developed an on-slide immunostaining method. We used the cell-specific neuropeptide markers anti-RYamide (INNOS) and anti-RGWa (INRGW) [@Conzelmann2011] to mark the interneurons. Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 4---figure supplement 2). In both wild type and *NOS*-mutant larvae, INNOS cells showed increase calcium signals during the inhibition phase of the cPRCs. In contrast, calcium signals increased in INRGW cells in wild-type larvae during the second activation phase of cPRCs, a response that was lacking in *NOS*-mutants. This suggests that there is no circuit output to the Ser-h1 cells and the ciliary band in *NOS*-mutants, consistent with the observed behavioural defects.

![**Figure 4. NO produced by UV stimulation to cPRCs induces a depolarising response in cPRC via NIT-GC1.** (A) The intensity of DAF-FM in INNOS changes over time when cPRC (top) and other cells (NC) (bottom) are stimulated with UV (405 nm). Hereafter, light purple squares indicate the timing of UV stimulation, gray traces show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark lines and the bands show the mean value and the standard error range of the mean. (B) Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in WT (purple) and NOS mutant (NOSΔ23) (gray) larvae. (C) Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in control (non-injection) and NIT-GC1 knockdown by morpholino (NITGC1-MO1 and 2) larvae. (D) Co-expression analysis image of the cPRC marker gene (NIT-GC1: magenta) and the INNOS marker gene (RYa-pNP: green) using in situ HCR. White arrowheads indicate INNOS strongly expressing RYa-pNP. (E) INNOS (blue dashed line) identified from calcium imaging using GCaMP6s by their position in relation to the cPRC (purple dashed line). (F) Comparison of changes over time in GCaMP6s intensities in cPRC (purple) and INNOS (blue) when cPRC is stimulated with UV in WT (left) and NOS mutant (NOSΔ11: right) larvae. (G) Comparison of changes over time in GCaMP6s intensities in cPRC (purple) and INRGW (green) when the cPRC are stimulated with UV in WT (left) and NOS mutant (NOSΔ23: right) larvae. (H)(I)(J)](../figures/Fig4.png){width="100%"}
